Higher-Order unification for free!

### 2024-04-12 paper/main.tex
sec 2.3 with spellchecker

\subsection{The problem: logic-program simulation}
We represent a logic program \emph{run} in \Fo{} as
a sequence of \emph{steps} of length $\mathcal{N}$.
At each step $p$ we unify two terms, $\foUnifPb_{p_l}$ and
$\foUnifPb_{p_r}$, taken from the list of all unification
problems \foUnifPb.
% \footnote{If the same rule is used multiple times in a run we
% consider as many copies as needed of the terms composing the
% rules, with fresh unification variables each time. Also, we do not
% model backtracking directly, but rather the fact that a step in a run may
% fail, and in that case the entire run is said to fail. A regular logic program
% that succeeds is modeled as a many runs, with possibly overlapping starts,
% where at least one run does not end in a failure. }
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$, that is the result of the
logic-program execution.
%
\newcommand{\C}[4]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3,#4)}
\newcommand{\D}[4]{\ensuremath{\langle #1,#2,#3 \rangle^{-1}\mapsto #4}}
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[4]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
%
$$
\begin{array}{l}
\stepF{\foUnifPb}{p}{\rho}{\rho'}
\deff
\stepFD{\foUnifPb}{p}{\rho}{\rho'}\vspace{2pt}\\
\runF{\foUnifPb}{\mathcal{N}}{\rho}
\deff
\runFD{\foUnifPb}{\mathcal{N}}
\end{array}
$$

\noindent
\todo{say that backtracking is not important} 
In order to simulate a \Fo{} logic program in \Ho{} we compile
each \Fo{} term $s$ in \foUnifPb to a \Ho{} term $t$.
We write this translation $\C{s}{t}{m}{l}$. The implementation of the compiler
is detailed in \cref{sec:simulation,sec:eta,sec:beta}, here we just point
out that it additionally produce a variable mapping $m$ and list of links $l$.
The variable map connects unification variables in \Ho to variables
in \Fo{} and is used to ``decompile'' the assignment,
$\D{\sigma}{m}{l}{\rho}$. Links are an accessory piece of information whose
description is deferred to \cref{sec:nutshell}.
We write $\hoUnifPb_p = \{~ \hoUnifPb_{p_l}, \hoUnifPb_{p_r} ~\}$
and $s \in \foUnifPb \Leftrightarrow \exists p, s \in \foUnifPb_p$.

@@

\noindent
By analogy with \foUnifPb, we write $\hoUnifPb_{p_l}$ and $\hoUnifPb_{p_r}$
for the two \Ho{} terms being unified at step $p$, and we write $\hoUnifPb_p$
for the set $\{~ \hoUnifPb_{p_l}, \hoUnifPb_{p_r} ~\}$.\\
\hstep{} is made of two sub-steps: a call to the meta language
unification and a check for \progress{} on the set of links, that intuitively
will compensate for the weaker equational theory honoured by \Ue.
\hrun{} compiles all terms in \foUnifPb{}, then executes each step and
finally decompiles the solution.
We claim:

@@

\begin{proposition}[Simulation fidelity]\label{prop:fidelity}
In the context of$\;$ \hrun, if $~\foUnifPb \subseteq \llambda$ we have that
$\forall p \in 1 \ldots \mathcal{N},$
$$
\stepF{\foUnifPb}{p}{\rho_{p-1}}{\rho_p}
\Leftrightarrow
\stepH{\hoUnifPb}{p}{\sigma_{p-1}}{\linkStore_{p-1}}{\sigma_p}{\linkStore_p}
$$
\end{proposition}
\noindent
In particular this property guarantees that a \emph{failure} in the \Fo{} run
is matched by a failure in \Ho{} \emph{at the same step}. We consider this
property very important from a practical point of view since it guarantees
that the execution traces are strongly related and in turn this enables a user
to debug a logic program in \Fo{} by looking at its execution trace in
\Ho{}.

@@

In other words if the two terms involved in a step
re-enter \llambda, then \hstep and \fstep are again related, even if
$\foUnifPb \not\subseteq \llambda$ and hence \cref{prop:fidelity} does not apply.
Indeed, the main difference between \cref{prop:fidelity} and \cref{prop:fidelity-recovery}
is that the assumption of the former is purely static, it can be checked upfront. 
When this assumption is not satisfied one can still simulate a logic program and
have guarantees of fidelity if, at run time, decidability of higher-order
unification is restored.

This property has a practical relevance since in many logic programming
implementations, including Elpi, the order in which unification problems
are tackled does matter.
The simplest example is the sequence $F \UnifRel \lambda x.a$ and
$F \appsep a \UnifRel a$: the second problem is not in \llambda and has two
unifiers, namely $\sigma_1 = \{~ F \mapsto \lambda x.x ~\}$ and
$\sigma_2 = \{~ F \mapsto \lambda x.a ~\}$. The first problem picks $\sigma_2$
making the second problem re-enter \llambda.

\paragraph{Backtracking} We omit it from our model of a logic programs execution
since it pays a very minor role, orthogonal to higher-order unification.
We point out that each \emph{run} corresponds to a (proof search) branch in the
logic program that either fails at some point, or succeeds. A computation that
succeeds by backtracking, exploring multiple branches, could be
modeled as set of runs with (possibly non empty) common prefixes.

>>>>

\subsection{The problem: logic-program simulation}
We represent a logic program \emph{run} in \Fo{} as
a sequence of \emph{steps} of length $\mathcal{N}$.
At each step, $p$ we unify two terms, $\foUnifPb_{p_l}$ and
$\foUnifPb_{p_r}$, taken from the list of all unification
problems \foUnifPb.
% \footnote{If the same rule is used multiple times in a run we
% consider as many copies as needed of the terms composing the
% rules, with fresh unification variables each time. Also, we do not
% model backtracking directly, but rather the fact that a step in a run may
% fail, and in that case the entire run is said to fail. A regular logic program
% that succeeds is modeled as a many runs, with possibly overlapping starts,
% where at least one run does not end in a failure. }
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$, which is the result of the
logic program execution.
%
\newcommand{\C}[4]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3,#4)}
\newcommand{\D}[4]{\ensuremath{\langle #1,#2,#3 \rangle^{-1}\mapsto #4}}
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[4]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
%
$$
\begin{array}{l}
\stepF{\foUnifPb}{p}{\rho}{\rho'}
\deff
\stepFD{\foUnifPb}{p}{\rho}{\rho'}\vspace{2pt}\\
\runF{\foUnifPb}{\mathcal{N}}{\rho}
\deff
\runFD{\foUnifPb}{\mathcal{N}}
\end{array}
$$

\noindent
In order to simulate a \Fo{} logic program in \Ho{}, we compile
each \Fo{} term $s$ in \foUnifPb to a \Ho{} term $t$.
We write this translation as $\C{s}{t}{m}{l}$. The implementation of the compiler
is detailed in \cref{sec:simulation,sec:eta,sec:beta}, here we just point
out that it additionally produces a variable mapping $m$ and a list of links $l$.
The variable map connects unification variables in \Ho to variables
in \Fo{} and is used to ``decompile'' the assignment,
$\D{\sigma}{m}{l}{\rho}$. Links are an accessory piece of information whose
description is deferred to \cref{sec:nutshell}.
We write $\hoUnifPb_p = \{~ \hoUnifPb_{p_l}, \hoUnifPb_{p_r} ~\}$
and $s \in \foUnifPb \Leftrightarrow \exists p, s \in \foUnifPb_p$.

@@

\noindent
By analogy with \foUnifPb, we write $\hoUnifPb_{p_l}$ and $\hoUnifPb_{p_r}$
for the two \Ho{} terms being unified at step $p$, and we write $\hoUnifPb_p$
for the set $\{~ \hoUnifPb_{p_l}, \hoUnifPb_{p_r} ~\}$.\\
\hstep{} is made of two sub-steps: a call to the meta language
unification and a check for \progress{} on the set of links, that intuitively
will compensate for the weaker equational theory honored by \Ue.
\hrun{} compiles all terms in \foUnifPb{}, then executes each step, and
finally decompiles the solution.
We claim:

@@

\begin{proposition}[Simulation fidelity]\label{prop:fidelity}
In the context of$\;$ \hrun, if $~\foUnifPb \subseteq \llambda$ we have that
$\forall p \in 1 \ldots \mathcal{N},$
$$
\stepF{\foUnifPb}{p}{\rho_{p-1}}{\rho_p}
\Leftrightarrow
\stepH{\hoUnifPb}{p}{\sigma_{p-1}}{\linkStore_{p-1}}{\sigma_p}{\linkStore_p}
$$
\end{proposition}
\noindent
In particular, this property guarantees that a \emph{failure} in the \Fo{} run
is matched by a failure in \Ho{} \emph{at the same step}. We consider this
property very important from a practical point of view since it guarantees
that the execution traces are strongly related, and in turn, this enables a user
to debug a logic program in \Fo{} by looking at its execution trace in
\Ho{}.

@@

In other words, if the two terms involved in a step
re-enter \llambda, then \hstep and \fstep are again related, even if
$\foUnifPb \not\subseteq \llambda$ and hence \cref{prop:fidelity} does not apply.
Indeed, the main difference between \cref{prop:fidelity} and \cref{prop:fidelity-recovery}
is that the assumption of the former is purely static, it can be checked upfront. 
When this assumption is not satisfied, one can still simulate a logic program and
have guarantees of fidelity if, at run time, decidability of higher-order
unification is restored.

This property has practical relevance since in many logic programming
implementations, including Elpi, the order in which unification problems
are tackled does matter.
The simplest example is the sequence $F \UnifRel \lambda x.a$ and
$F \appsep a \UnifRel a$: the second problem is not in \llambda and has two
unifiers, namely $\sigma_1 = \{~ F \mapsto \lambda x.x ~\}$ and
$\sigma_2 = \{~ F \mapsto \lambda x.a ~\}$. The first problem picks $\sigma_2$,
making the second problem re-enter \llambda.

\paragraph{Backtracking} We omit it from our model of a logic program's execution
since it plays a very minor role, orthogonal to higher-order unification.
We point out that each \emph{run} corresponds to a (proof search) branch in the
logic program that either fails at some point, or succeeds. A computation that
succeeds by backtracking, exploring multiple branches, could be
modeled as a set of runs with (possibly non-empty) common prefixes.
